Scipion Tailored for TEM Facilities: Image Processing in Streaming

### 2018-02-28 2.overall.tex
change functioning by execution
Update 2.overall.tex

\section{Main Modifications introduced in  \scipion: Streaming, Monitors and Consensus}
\label{overall}

In the following we describe the main modifications made to \scipion in order to facilitate its deployment and functioning in centralized facilities.

The main change introduced in \scipion is \emph{stream processing}. Stream processing has been designed to analyze and react on real-time so that enables analysis of data as it is produced.
In this way a new level of parallelization has been reached since the  data generated by the microscope continuously flows through a chain of protocols. For example, as movies are imported to the system they are aligned and their CTF is estimated. In \scipion version 1.1, users may find  the first steps in the image processing adapted to streaming, including movie alignment and CTF estimation. We have also implemented particle picking and extraction, that are going to be released soon in the new version 1.2. In developers version, several 2D classification algorithms are being adapted for this new way of processing.

\scipion stream implementation is based on periodic updates of the protocol input and output data during the image processing. 
[**JMRT The following description is not accurate...Since \scipion runs on threads, it is possible to devote one thread to update these information while the rest of the CPU power is devoted to run the different protocols. The use of threads marks an important difference between running a script or using \scipion, since  each module is run independently creating a very flexible processing framework]. [**JMRT The following sentence seems too colloquial compared with the other style...For example, do you want to add a new protocol in your workflow while it is running? No problem, do it]. \scipion allows users to modify  workflows on the fly without modifying a single code line. The modification will not interfere with the already executed steps and may take advantage from all already computed data. Did something go wrong and the microscope stopped? By default \scipion will stop after 24 hours with no microscope activity, but if you wish to continue your acquisition after a stop just press the continue button, there is no need to restart from the beginning.

@@

Since \scipion original publication we have developed many \emph{metaprotocols} [**JMRT not sure if the 
introduction of the metaprotocols definition helps the reader...anyway, there are metaprotocols in the first
version, so this is not completely new], that is, 
protocols that either check the progress of other protocols or compare the results of equivalent protocols.

@@

The second set of metaprotocols are grouped under the name \emph{consensus} [**JMRT Consesus protocol existed in the first release]. For a given logical step (for example, particle picking) these protocols check if the datasets obtained from the same input data using a different algorithm are consistent. Continuing with the particle picking example, \emph{consensus picking} will compare the particles selected for each one of the executed particle-picking algorithms and produce as output a particle set containing those particles selected by most of the algorithms.

Despite the modifications introduced to \scipion data and process model for stream processing, the framework still provides full data provenance. Data provenance is an important form of metadata that describes how a particular data set was generated by detailing: the steps, input data and parameters used in the computation. Data provenance guaranties reproducibility and traceability. %Since provenance is solved by Scipion and not Scipionbox we will not discuss here how it is been achieved. We just note that, before choosing which data has to be stored, it is necessary to define how these data have to be structured so that they can be later recovered and understood . %This abstract modeling of the data also simplifies the software interoperability offered by Scipion.

@@

The workflow starts two independent protocols in parallel. The first protocol estimates the camera gain at each pixel for a few  movies \citep{sorzano2018}. % In the CNB setup the microscope camera is calibrated and internally performs the gain correction before writing the movies so the estimated gain image should be almost constant and equal to 1. 
The output of this protocol serves as verification of the validity of the experimental gain image for the whole data set. The second protocol performs direct detector movie alignment. Currently \scipion offers  four  methods for movie alignment: motioncorr2 (Li et al., 2013), correlation (an unpublished CPU version of the motioncorr2 method developed in Xmipp [**JMRT this one is only global-alignment, so I will not say that is CPU version of motioncorr2), unblur (Grant and Grigorieff, 2015) and optical flow (Abrishami et al., 2015). Within \scipion, movie frames can be aligned using any one of these approaches or any combination. Following the standard [**JMRT standard? recommend or not recommend? maybe common?] processing workflow, the resulting micrographs are screened to eliminate those with excessive astigmatism or drift [**JMRT This screen is not done after CTF?]. Scipion provides three CTF estimation methods:
CTFFIND4 (Rohou and Grigorieff, 2015), gCTF (Zhang, 2016) and Xmipp CTF estimation (Sorzano et al., 2007; Vargas et al., 2013). At this point the user may compute the CTFs using several algorithms and compare them with the protocol \protocol{CTF consensus} that report up to which frequency two estimation of the same micrograph are equivalent. Further automatic filtering based on CTF is available through the protocol \protocol{CTF filtering} that selects micrographs based on the astigmatism, resolution and defocus. Another criteria for CTF selection is iciness, that is,  the ratio of the intensity in the resolution band between 3.5 and 3.9 \AA, and between 6 and 30 \AA, which gives a good estimate of the ice crystal content of the vitreous specimen.

Up to this point the image processing pipeline is quite solid and may be executed in streaming without user intervention. The next step is to select (pick) particles. We have recently adapted eight particle picking methods in \scipion  to streaming. In this way,  particles may be selected as soon as any new micrograph is available. Some picking algorithms require a training step in which either user intervention or information from a previous selection is needed. The streaming functionality is separated from the picking logic, so if a new algorithm is added in the future it will work in streaming without extra effort from the develoepr. For those users that execute several picking algorithms, the protocol \protocol{picking consensus} may be useful. \protocol{picking consensus}  calculates the agreement between different particle-picking algorithms and produces a list with the particle selected by most of them.

@@

The proposed pipeline is very flexible to decide which particular algorithms will be used by a given user. At the CNB (and collaborating institutions) we collect protocol usage statistics and sort the different algorithms by popularity.  This information is available at  \url{http://scipion.i2pc.es/report_protocols/protocolTable/} and may help users to select between the different algorithms.

>>>>

\section{Main Modifications introduced in  \scipion: Streaming, Monitors and Consensus}
\label{overall}
[**ROB this section goes not flow smoothly, needs a second pass]

In the following we describe the main modifications made to \scipion in order to facilitate its deployment and execution in centralized facilities.

The main change introduced in \scipion is \emph{stream processing}. Stream processing has been designed to analyze and react on real-time so that enables analysis of data as it is produced.
In this way, a new level of parallelization has been reached since the  data generated by the microscope continuously flows through a chain of protocols. For example, as movies are imported to the system they are aligned and their CTF is estimated. In \scipion version 1.1, users may find  the first steps in the image processing adapted to streaming, including movie alignment and CTF estimation. We have also implemented particle picking and extraction, that are going to be released soon in the new version 1.2. In the developers version, several 2D classification algorithms are being adapted for this new way of processing.

\scipion stream implementation is based on periodic updates of the protocol input and output data during the image processing. 
[**JMRT The following description is not accurate...Since \scipion runs on threads, it is possible to devote one thread to update these information while the rest of the CPU power is devoted to run the different protocols. The use of threads marks an important difference between running a script or using \scipion, since  each module is run independently creating a very flexible processing framework][**ROB: Any suggestion to make it more accurate?]. \scipion allows users to modify the workflow on the fly without modifying a single code line. The modification will not interfere with the already executed steps and may take advantage from all already computed data. By default \scipion will automatically stop after 24 hours with no microscope activity, but you may  continue the acquisition after a stop by pressing the continue button, there is no need to restart from the beginning.

@@

Since \scipion original publication we have developed many \emph{metaprotocols} [**JMRT not sure if the 
introduction of the metaprotocols definition helps the reader...anyway, there are metaprotocols in the first
version, so this is not completely new][**ROB there is no claim that the concept of metaprotocol is new, The only claim is that many protocols of this kind have been introduced in this version], that is, 
protocols that either check the progress of other protocols or compare the results of equivalent protocols.

@@

The second set of metaprotocols are grouped under the name \emph{consensus} [**JMRT Consesus protocol existed in the first release, **ROB Ok this is true I need to think about how to rewrite this]. For a given logical step (for example, particle picking) these protocols check if the datasets obtained from the same input data using a different algorithm are consistent. Continuing with the particle picking example, \emph{consensus picking} will compare the particles selected for each one of the executed particle-picking algorithms and produce as output a particle set containing those particles selected by most of the algorithms.

Despite the modifications introduced to \scipion data and process model, the framework still provides full data provenance. Data provenance is an important form of metadata that describes how a particular data set was generated by detailing: the steps, input data and parameters used in the computation. Data provenance guaranties reproducibility and traceability. %Since provenance is solved by Scipion and not Scipionbox we will not discuss here how it is been achieved. We just note that, before choosing which data has to be stored, it is necessary to define how these data have to be structured so that they can be later recovered and understood . %This abstract modeling of the data also simplifies the software interoperability offered by Scipion.

@@

The workflow starts two independent protocols in parallel. The first protocol estimates the camera gain at each pixel for a few  movies \citep{sorzano2018}. % In the CNB setup the microscope camera is calibrated and internally performs the gain correction before writing the movies so the estimated gain image should be almost constant and equal to 1. 
The output of this protocol serves as verification of the validity of the experimental gain image for the whole data set. The second protocol performs direct detector movie alignment. Currently \scipion offers  four  methods for movie alignment: motioncorr2 (Li et al., 2013), correlation (an unpublished CPU version that implements the global aligment method used by motioncorr2), unblur (Grant and Grigorieff, 2015) and optical flow (Abrishami et al., 2015). Then the CTF of the resulting micrographs are calculated. Scipion provides three CTF estimation methods:
CTFFIND4 (Rohou and Grigorieff, 2015), gCTF (Zhang, 2016) and Xmipp CTF estimation (Sorzano et al., 2007; Vargas et al., 2013). At this point the user may compute the CTFs using several algorithms and compare them with the protocol \protocol{CTF consensus} that reports up to which frequency two estimation of the same micrograph are equivalent. Further automatic filtering based on CTF is available through the protocol \protocol{CTF filtering} that selects micrographs based on the astigmatism, resolution and defocus. Another criteria for CTF selection is iciness, that is,  the ratio of the intensity in the resolution band between 3.5 and 3.9 \AA, and between 6 and 30 \AA, which gives a good estimate of the ice crystal content of the vitreous specimen.

Up to this point the image processing pipeline is quite solid and may be executed in streaming without user intervention. The next step is to select (pick) particles. We have recently adapted eight particle picking methods in \scipion  to streaming. In this way,  particles may be selected as soon as any new micrograph is available. Some picking algorithms require a training step in which either user intervention or information from a previous selection is needed. The streaming functionality is separated from the picking logic, so if in the future a new algorithm is added, it will work in streaming without extra effort from the algorithm developer. For those users that execute several picking algorithms, the protocol \protocol{picking consensus} may be useful. \protocol{picking consensus}  calculates the agreement between different particle-picking algorithms and produces a list with the particle selected by most of them.

@@

The proposed pipeline is very flexible and can be easily modified by each user. At the CNB (and collaborating institutions) we collect protocol usage statistics and sort the different algorithms by popularity.  This information is available at  \url{http://scipion.i2pc.es/report_protocols/protocolTable/} and may help users to select between the different algorithms.
